---
title: "IMDB Sentiment Analysis (MLP)"
subtitle: "Biostat 203B"
author: "Dr. Hua Zhou @ UCLA"
date: today
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
---

## Set up

This example tries to reproduce the two-layer MLP for classifying IMDB reviews based on bags-of-words.

Display system information for reproducibility.

:::  {.panel-tabset}

#### Python

```{python}
import IPython
print(IPython.sys_info())
```

#### R

```{r}
sessionInfo()
```

:::

Load libraries.

::: {.panel-tabset}

#### Python

```{python}
# Numpy
import numpy as np
# Plotting tool
import matplotlib.pyplot as plt
# Load Tensorflow and Keras
import tensorflow as tf
from tensorflow import keras
from tensorflow.keras import layers
```

#### R

```{r}
library(keras)
```

:::


## Prepare data

From documentation:  

> Dataset of 25,000 movies reviews from IMDB, labeled by sentiment (positive/negative). Reviews have been preprocessed, and each review is encoded as a sequence of word indexes (integers). For convenience, words are indexed by overall frequency in the dataset, so that for instance the integer "3" encodes the 3rd most frequent word in the data. This allows for quick filtering operations such as: "only consider the top 10,000 most common words, but eliminate the top 20 most common words".

Retrieve IMDB data:

::: {.panel-tabset}

#### Python

```{python}
max_features = 10000 # to be consistent with lasso example
batch_size = 32

print('Loading data...')
(x_train, y_train), (x_test, y_test) = keras.datasets.imdb.load_data(
  num_words = max_features
  )
```

Sizes of training and test sets:
```{python}
print(len(x_train), 'train sequences')
print(len(x_test), 'test sequences')
```

#### R

```{r}
max_features <- 10000 # to be consistent with lasso example

cat('Loading data...\n')
imdb <- dataset_imdb(num_words = max_features)
imdb$train$x[[1]]
imdb$train$y[[1]]
```

Sizes of training and test sets:
```{r}
x_train <- imdb$train$x
y_train <- imdb$train$y
x_test <- imdb$test$x
y_test <- imdb$test$y

cat(length(x_train), 'train sequences\n')
cat(length(x_test), 'test sequences\n')
```

:::

Create the bag of words matrices.

::: {.panel-tabset}

#### Python

```{python}
from scipy import sparse

def one_hot(sequences, dimension):
  seqlen = [len(sequences[i]) for i in range(len(sequences))]
  n = len(seqlen)
  rowind = np.repeat(range(n), seqlen)
  colind = np.concatenate(sequences)
  vals = np.ones(len(rowind))
  # Has to be CSR format for batching; CSC doesn't work for Keras
  return sparse.coo_matrix((vals, (rowind, colind)), shape = (n, dimension)).tocsr()

# Train
x_train_1h = one_hot(x_train, max_features)
x_train_1h.shape
# Sparsity of train set
x_train_1h.nnz / np.prod(x_train_1h.shape)
# Test
x_test_1h = one_hot(x_test, max_features)
x_test_1h.shape
# Sparsity of test set
x_test_1h.nnz / np.prod(x_test_1h.shape)
```
#### R

```{r}
library(Matrix)

one_hot <- function(sequences, dimension) {
  seqlen <- sapply(sequences, length)
  n <- length(seqlen)
  rowind <- rep(1:n, seqlen)
  colind <- unlist(sequences)
  sparseMatrix(
    i = rowind,
    j = colind,
    dims = c(n, dimension)
  )
}

# Train
x_train_1h <- one_hot(x_train, max_features)
dim(x_train_1h)
# Proportion of nonzeros
nnzero(x_train_1h) / (25000 * max_features)
# Test
x_test_1h <- one_hot(x_test, max_features)
dim(x_test_1h)
# Proportion of nonzeros
nnzero(x_test_1h) / (25000 * max_features)
```

:::

Encode $y$ as binary class matrix:

::: {.panel-tabset}

#### Python

```{python}
y_train = keras.utils.to_categorical(y_train, 2)
y_test = keras.utils.to_categorical(y_test, 2)
# Train
y_train.shape
# Test
y_test.shape
```

#### R

```{r}
y_train <- to_categorical(y_train, 2)
y_test <- to_categorical(y_test, 2)
# Train
dim(y_train)
# Test
dim(y_test)
```

:::

## Build model

::: {.panel-tabset}

#### Python

```{python}
model = keras.Sequential([
  keras.Input(shape = (max_features,)),
  layers.Dense(units = 16, activation = 'ReLU'),
  layers.Dense(units = 16, activation = 'ReLU'),
  layers.Dense(units = 2, activation = 'softmax')
])

model.summary()
```

Compile model:
```{python}
# try using different optimizers and different optimizer configs
model.compile(
  loss = 'binary_crossentropy',
  optimizer = 'adam',
  metrics = ['accuracy']
)
```

#### R

```{r}
model <- keras_model_sequential() %>%
  layer_dense(units = 16, activation = "ReLU", input_shape = max_features) %>% 
  layer_dense(units = 16, activation = "ReLU") %>% 
  layer_dense(units = 2, activation = 'softmax')

# Try using different optimizers and different optimizer configs
model %>% compile(
  loss = 'binary_crossentropy',
  optimizer = 'adam',
  metrics = c('accuracy')
)
summary(model)
```

:::

## Training

::: {.panel-tabset}

#### Python

```{python}
print('Train...')
history = model.fit(
  x_train_1h, y_train,
  batch_size = batch_size,
  epochs = 20,
  validation_data = (x_test_1h, y_test), 
  verbose = 2 # one line per epoch
)
```

Visualize training process:
```{python}
plt.figure()
plt.ylabel("Loss (training and validation)")
plt.xlabel("Training Steps")
plt.ylim([0, 2])
plt.plot(history.history["loss"])
plt.plot(history.history["val_loss"])
plt.show()

plt.figure()
plt.ylabel("Accuracy (training and validation)")
plt.xlabel("Training Steps")
plt.ylim([0, 1])
plt.plot(history.history["accuracy"])
plt.plot(history.history["val_accuracy"])
plt.show()
```

#### R

```{r}
batch_size <- 32

cat('Train...\n')
system.time({
history <- model %>% fit(
  x_train_1h, y_train,
  batch_size = batch_size,
  epochs = 20,
  validation_data = list(x_test_1h, y_test),
  verbose = 2
)
})
```

Visualize training process:
```{r}
plot(history)
```

:::

## Testing

::: {.panel-tabset}

#### Python

```{python}
score, acc = model.evaluate(
  x_test_1h, y_test,
  batch_size = batch_size,
  verbose = 2
)
print('Test score:', score)
print('Test accuracy:', acc)
```

#### R

```{r}
scores <- model %>% evaluate(
  x_test_1h, y_test,
  batch_size = batch_size
)
```

```{r}
cat('Test score:', scores[[1]])
cat('Test accuracy', scores[[2]])
```

:::