---
title: "IMDB Sentiment Analysis (Warm-Start From Pretrained Embedding)"
subtitle: "Biostat 203B"
author: "Dr. Hua Zhou @ UCLA"
date: today
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
---

## Setup

Display system information for reproducibility.

:::  {.panel-tabset}

#### Python

```{python}
import IPython
print(IPython.sys_info())
```

#### R

```{r}
sessionInfo()
```

:::

Load libraries.

::: {.panel-tabset}

#### Python

```{python}
# Plotting tool
import matplotlib.pyplot as plt
# Load Tensorflow and Keras
import tensorflow as tf
from tensorflow import keras
from tensorflow.keras import layers
import tensorflow_hub as hub
```

#### R

```{r}
library(keras)
library(tfhub)
```

:::

Source: <https://tensorflow.rstudio.com/tutorials/keras/text_classification_with_hub>

## Prepare data

Different from earlier experiment of fitting LSTM on IMDB data, we will start from the original raw text of IMDB reviews. 

We download the IMDB dataset from a static url (if it's not already in the cache):

::: {.panel-tabset}

#### R

```{r}
if (dir.exists("aclImdb/"))
  unlink("aclImdb/", recursive = TRUE)
url <- "https://ai.stanford.edu/~amaas/data/sentiment/aclImdb_v1.tar.gz"
dataset <- get_file(
  "aclImdb_v1",
  url,
  untar = TRUE,
  cache_dir = '.',
  cache_subdir = ''
)
unlink("aclImdb/train/unsup/", recursive = TRUE)
```

We can then create a TensorFlow dataset from the directory structure using the `text_dataset_from_directory` function:

::: {.panel-tabset}

#### Python

```{python}
batch_size = 512
seed = 425

train_data = keras.utils.text_dataset_from_directory(
  'aclImdb/train',
  batch_size = batch_size,
  validation_split = 0.2,
  subset = 'training',
  seed = seed
)
validation_data = keras.utils.text_dataset_from_directory(
  'aclImdb/train',
  batch_size = batch_size,
  validation_split = 0.2,
  subset = 'validation',
  seed = seed
)
test_data = keras.utils.text_dataset_from_directory(
  'aclImdb/test',
  batch_size = batch_size
)
```

#### R

```{r}
batch_size <- 512
seed <- 425

train_data <- text_dataset_from_directory(
  'aclImdb/train',
  batch_size = batch_size,
  validation_split = 0.2,
  subset = 'training',
  seed = seed
)
validation_data <- text_dataset_from_directory(
  'aclImdb/train',
  batch_size = batch_size,
  validation_split = 0.2,
  subset = 'validation',
  seed = seed
)
test_data <- text_dataset_from_directory(
  'aclImdb/test',
  batch_size = batch_size
)
```
:::

Let’s take a moment to understand the format of the data. Each example is a sentence representing the movie review and a corresponding label. The sentence is not preprocessed in any way. The label is an integer value of either 0 or 1, where 0 is a negative review, and 1 is a positive review.

Let’s print first 10 examples.

::: {.panel-tabset}

#### Python

```{python}
batch = list(train_data.as_numpy_iterator())[0]
batch[0][0]
```
First 10 labels:
```{python}
batch[1][0:9]
```

#### R

```{r}
batch <- train_data %>%
  reticulate::as_iterator() %>%
  reticulate::iter_next()

batch[[1]][1]
```
Let’s also print the first 10 labels.
```{r}
batch[[2]][1:10]
```

:::

## Build model

Let’s first create a Keras layer that uses a TensorFlow Hub model to embed the sentences, and try it out on a couple of input examples. Note that no matter the length of the input text, the output shape of the embeddings is: `(num_examples, embedding_dimension)`.

::: {.panel-tabset}

#### Python

```{python}
embedding = "https://tfhub.dev/google/nnlm-en-dim50/2"
hub_layer = hub.KerasLayer(
  handle = embedding, 
  # Enable fine-tuning (takes longer)
  trainable = True
  )
# Embed the first training texts
hub_layer(batch[0][0:1])
```

#### R

```{r}
embedding <- "https://tfhub.dev/google/nnlm-en-dim50/2"
hub_layer <- tfhub::layer_hub(
  handle = embedding, 
  # Enable fine-tuning (takes longer)
  trainable = TRUE
  )
# Embed the first training texts
hub_layer(batch[[1]][1:2])
```

:::

Let’s now build and compile the full model:

::: {.panel-tabset}

#### Python

```{python}
model = keras.Sequential([
  hub_layer,
  layers.Dense(units = 16, activation = 'relu'),
  layers.Dense(units = 1, activation = 'sigmoid')
]
)
```

```{python}
model.compile(
  optimizer = 'adam',
  loss = "binary_crossentropy",
  metrics = 'accuracy'  
)
```


#### R

```{r}
model <- keras_model_sequential() %>%
  hub_layer() %>%
  layer_dense(16, activation = 'relu') %>%
  layer_dense(1)
```

Compile model:
```{r}
model %>% compile(
  optimizer = 'adam',
  loss = loss_binary_crossentropy(from_logits = TRUE),
  metrics = 'accuracy'
)
```

:::

:::

## Training

::: {.panel-tabset}

#### Python

```{python}
history = model.fit(
  train_data,
  epochs = 10,
  validation_data = validation_data,
  verbose = 2
)

model.summary()
```
Visualize training process:
```{python}
plt.figure()
plt.ylabel("Loss (training and validation)")
plt.xlabel("Training Epoches")
plt.ylim([0, 1])
plt.plot(history.history["loss"])
plt.plot(history.history["val_loss"])
plt.show()

plt.figure()
plt.ylabel("Accuracy (training and validation)")
plt.xlabel("Training Epoches")
plt.ylim([0, 1])
plt.plot(history.history["accuracy"])
plt.plot(history.history["val_accuracy"])
plt.show()
```

#### R

```{r}
system.time({
history <- model %>% fit(
  train_data,
  epochs = 10,
  validation_data = validation_data,
  verbose = 2
)
})

summary(model)
```

Visualize training process:
```{r}
plot(history)
```

:::

## Testing

::: {.panel-tabset}

#### Python

```{python}
results = model.evaluate(test_data, verbose = 2)
results
```

#### R

```{r}
results <- model %>% evaluate(test_data, verbose = 2)
results
```

:::