---
title: "IMDB Sentiment Analysis (Transformer)"
subtitle: "Econ 425T / Biostat 203B"
author: "Dr. Hua Zhou @ UCLA"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
engine: knitr
knitr:
  opts_chunk: 
    fig.align: 'center'
    # fig.width: 6
    # fig.height: 4
    message: FALSE
    cache: false
---

Source: <https://keras.io/examples/nlp/text_classification_with_transformer/>

Display system information for reproducibility.

:::  {.panel-tabset}

#### Python

```{python}
import IPython
print(IPython.sys_info())
```

#### R

```{r}
sessionInfo()
```

:::

Load libraries.

::: {.panel-tabset}

#### Python

```{python}
# Plotting tool
import matplotlib.pyplot as plt
# Load Tensorflow and Keras
import tensorflow as tf
from tensorflow import keras
from tensorflow.keras import layers
```

#### R

```{r}
library(keras)
```

:::

## Implement a transformer block as a layer

```{python}
class TransformerBlock(layers.Layer):
    def __init__(self, embed_dim, num_heads, ff_dim, rate = 0.1):
        super().__init__()
        self.att = layers.MultiHeadAttention(num_heads = num_heads, key_dim = embed_dim)
        self.ffn = keras.Sequential(
            [layers.Dense(ff_dim, activation = "relu"), layers.Dense(embed_dim),]
        )
        self.layernorm1 = layers.LayerNormalization(epsilon = 1e-6)
        self.layernorm2 = layers.LayerNormalization(epsilon = 1e-6)
        self.dropout1 = layers.Dropout(rate)
        self.dropout2 = layers.Dropout(rate)

    def call(self, inputs, training):
        attn_output = self.att(inputs, inputs)
        attn_output = self.dropout1(attn_output, training = training)
        out1 = self.layernorm1(inputs + attn_output)
        ffn_output = self.ffn(out1)
        ffn_output = self.dropout2(ffn_output, training=training)
        return self.layernorm2(out1 + ffn_output)
```

## Implement embedding layer

Two seperate embedding layers, one for tokens, one for token index (positions).
```{python}
class TokenAndPositionEmbedding(layers.Layer):
    def __init__(self, maxlen, vocab_size, embed_dim):
        super().__init__()
        self.token_emb = layers.Embedding(input_dim=vocab_size, output_dim=embed_dim)
        self.pos_emb = layers.Embedding(input_dim=maxlen, output_dim=embed_dim)

    def call(self, x):
        maxlen = tf.shape(x)[-1]
        positions = tf.range(start = 0, limit = maxlen, delta = 1)
        positions = self.pos_emb(positions)
        x = self.token_emb(x)
        return x + positions
```

## Download and prepare dataset

```{python}
vocab_size = 20000  # Only consider the top 20k words
maxlen = 200  # Only consider the first 200 words of each movie review
(x_train, y_train), (x_val, y_val) = keras.datasets.imdb.load_data(num_words = vocab_size)
print(len(x_train), "Training sequences")
print(len(x_val), "Validation sequences")
x_train = keras.preprocessing.sequence.pad_sequences(x_train, maxlen = maxlen)
x_val = keras.preprocessing.sequence.pad_sequences(x_val, maxlen = maxlen)
```

## Create classifier model using transformer layer

Transformer layer outputs one vector for each time step of our input sequence. Here, we take the mean across all time steps and use a feed forward network on top of it to classify text.
```{python}
embed_dim = 32  # Embedding size for each token
num_heads = 2  # Number of attention heads
ff_dim = 32  # Hidden layer size in feed forward network inside transformer

inputs = layers.Input(shape = (maxlen,))
embedding_layer = TokenAndPositionEmbedding(maxlen, vocab_size, embed_dim)
x = embedding_layer(inputs)
transformer_block = TransformerBlock(embed_dim, num_heads, ff_dim)
x = transformer_block(x)
x = layers.GlobalAveragePooling1D()(x)
x = layers.Dropout(0.1)(x)
x = layers.Dense(20, activation = "relu")(x)
x = layers.Dropout(0.1)(x)
outputs = layers.Dense(2, activation = "softmax")(x)

model = keras.Model(inputs=inputs, outputs=outputs)
```

## Train and evaluate

```{python}
model.compile(
  optimizer = "adam", 
  loss = "sparse_categorical_crossentropy", 
  metrics = ["accuracy"]
  )
history = model.fit(
    x_train, 
    y_train, 
    batch_size = 32, 
    epochs = 2, 
    validation_data = (x_val, y_val),
    verbose = 2
)
```
