---
title: "Working with Databases - Part II"
subtitle: Biostat 203B
author: "Dr. Hua Zhou @ UCLA"
date: today
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
knitr:
  opts_chunk: 
    fig.align: 'center'
    fig.width: 6
    fig.height: 4
    message: FALSE
    cache: false
---

## Introduction

In this lecture we will demonstrate:

- Import data from bigish csv files (MIMIC-IV).

- Deposit data into an SQLite database.

- Query SQLite database. 

- Transform in database and plot in R.

## Machine information

Display machine information for reproducibility.
```{r}
sessionInfo()
```
Load necessary R packages
```{r}
library(tidyverse)
library(DBI)
library(RSQLite)
library(dbplyr)
```

```{bash}
# display version of SQLite
sqlite3 --version
```

## CSV file

The `/mnt/mimiciv/1.0` folder on the teaching server contains the MIMIC-IV data. On my Mac, it's at `/Users/huazhou/Documents/Box Sync/MIMIC/mimic-iv-1.0`.

Code in this lecture assumes that the MIMIC-IV data is available at `~/mimic`.

Display content of MIMIC-IV data folder:
```{bash}
ls -l ~/mimic
```

## Read CSVs and deposit to an SQLite database

Here, we will import only one csv file `icustays.csv.gz` for demonstration purpose. Motivated students can write a Bash script for loading all MIMIC-IV data files into a SQLite database and contribute to <https://github.com/MIT-LCP/mimic-code>.

Create an empty database file `mimiciv.sqlite`:
```{bash}
# bash command
touch mimiciv.sqlite
```

Deposit the `icu/icustatys.csv.gz` file:
```{bash}
# bash command
# delete icustays table if exists
sqlite3 mimiciv.sqlite 'DROP TABLE IF EXISTS icustays;'
```
Create an empty `icustays` table with data types. Because SQLite does not support date-time data type (<https://www.sqlite.org/datatype3.html>), we store `intime` and `outtime` as TEXT.
```{bash}
# bash command
sqlite3 mimiciv.sqlite 'CREATE TABLE icustays (
  subject_id INTEGER,
  hadm_id INTEGER,
  stay_id INTEGER,
  first_careunit TEXT,
  last_careunit TEXT,
  intime TEXT,
  outtime TEXT,
  los REAL
  )'
```

```{bash}
# bash command
zcat < ~/mimic/icu/icustays.csv.gz | \
  tail -n +2 | \
  sqlite3 mimiciv.sqlite -csv '.import /dev/stdin icustays'
```

## Read data from database

Connect to the database `mimiciii.sqlite` and list the tables:
```{r}
con <- dbConnect(
  RSQLite::SQLite(), 
  dbname = "./mimiciv.sqlite"
)
dbListTables(con)
```

Read the table `icustays`:
```{r}
icustays_tble <- tbl(con, "icustays") %>% 
  print(width = Inf)
```
How many rows?
```{r}
icustays_tble %>% 
  show_query() %>%
  summarise(n = n())
```

## Use dplyr with SQLite

Keep the first ICU stay for each patient:
```{r}
icustays_subset <- icustays_tble %>%
  # first ICU stay of each unique `subject_id`
  group_by(subject_id) %>%
  slice_min(intime) %>%
  ungroup() %>%
  # arrange(intime, .by_group = TRUE) %>%
  # slice_head(n = 1) %>%
  # left_join(icustays_tble, by = c("subject_id", "intime")) %>%
  # show_query() %>%
  print(width = Inf)
```

How many rows in `icustays_subset`?
```{r}
icustays_subset %>%
  show_query() %>%
  summarise(n = n())
```

## SQL query

`show_query` usefully shows the SQL query translated from dplyr query.
```{r}
class(icustays_subset)
show_query(icustays_subset)
```

## Transform in database, plot in R

```{r}
icustays_tble %>%
  group_by(subject_id) %>%
  summarise(n = n()) %>%
  ggplot() +
  geom_bar(mapping = aes(x = n)) + 
  labs(x = "# ICU stays of a patient")
```

## SQL translation

dbplyr package (a dplyr backend for databases) has a function, `translate_sql`, that lets you experiment with how R functions are translated to SQL:
```{r}
#| eval: false
translate_sql(x == 1 & (y < 2 | z > 3))
translate_sql(x ^ 2 < 10)
translate_sql(x %% 2 == 10)
translate_sql(paste(x, y))
translate_sql(mean(x))
translate_sql(mean(x, na.rm = TRUE))
```

## Timings

Let's compare the timings of dplyr (in-memory) and dbplyr (on disk database).

- dplyr using tibble:
```{r}
csvfile <- "~/mimic/icu/icustays.csv.gz"
icustays_tibble <- read_csv(csvfile)
timing_tibble <- system.time(
  icustays_tibble %>%
    group_by(subject_id) %>%
    summarize(n = n())
)
timing_tibble
```

- dbplyr using SQLite:
```{r}
icustays_sql <- tbl(con, "icustays")
timing_sql <- system.time(
  icustays_sql %>%
    group_by(subject_id) %>%
    summarize(n = n())
)
timing_sql
```

SQLite (`r timing_sql[3]` seconds) was much faster than tibble (`r timing_tibble[3]` seconds). But SQLite is disk-based, while the tibble is in memory. Why is the discrepancy?

## Laziness

dplyr/dbplyr uses lazy evaluation as much as possible, particularly when working with non-local backends.

- When building a query, often we don’t want the entire table. We want just enough to check if our query is working.

- Since we would prefer to run one complex query over many simple queries, laziness allows for verbs to be strung together.

- Therefore, by default dbplyr

    - won’t connect and query the database until absolutely necessary (e.g. show output),

    - and unless explicitly told to, will only query a handful of rows to give a sense of what the result will look like
    
```{r}
icustays_sql %>%
  group_by(subject_id) %>%
  summarize(n = n())
```

## Full query

To force a full query and return a complete table it is necessary to use the `collect` function.
```{r}
system.time(
  icustays_sql %>%
    group_by(subject_id) %>%
    summarize(n = n()) %>%
    collect()
)
```

## Old ggplot2 doesn't do ``Transform in database, plot in R"

Earlier we see dbplyr connects with ggplot2 (v3.3.5) seamlessly. Remember ggplot2 needs to `collect` the query results for plotting.
```{r}
icustays_sql %>%
  count(subject_id) %>%
  ggplot() + 
  geom_boxplot(mapping = aes(y = n))
```

Older version of ggplot2, e.g., v2.2.1, will output error. This is because ggplot2 needed to compute the count per bin by going through all the rows. But here `icustays_sql` is just a pointer to the SQLite table. We had to use the **transform in database, plot in R** strategy.
```{r}
icustays_sql %>%
  count(subject_id) %>%
  collect() %>%
  ggplot() + 
  geom_boxplot(mapping = aes(y = n))
```

## Close connection to database

```{r}
dbDisconnect(con)
```