---
title: "Dates and Times | r4ds Chapter 16"
subtitle: Biostat 203B
author: "Dr. Hua Zhou @ UCLA"
date: today
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
knitr:
  opts_chunk: 
    fig.align: 'center'
    fig.width: 6
    fig.height: 4
    message: FALSE
    cache: false
---

Display machine information for reproducibility.

::: {.panel-tabset}

#### R

```{r}
sessionInfo()
```

#### Python

```{python}
import IPython
print(IPython.sys_info())
```

#### Julia

```{julia}
using InteractiveUtils
versioninfo()
```

:::

Load tidyverse and lubridate (R), Pandas (Python), and DataFrames.jl (Julia).

::: {.panel-tabset}

#### R

```{r}
library(lubridate)
library(nycflights13)
library(tidyverse)
```

#### Python

```{python}
# Load the pandas library
import pandas as pd
# Load numpy for array manipulation
import numpy as np
```

#### Julia

```{julia}
using DataFrames, Pipe, StatsBase
```

:::


## Basics

Three types of data/time data:  

- **date**. Tibbles print it as `<date>`.  
- **time**. Tibbles print it as `<time>`.  
- **date-time**. Tibbles print it as `<dttm>`.  

In the `flights` tibble, the last variable `time_hour` is in the data-time format:

::: {.panel-tabset}

#### R

```{r}
flights |> print(width = Inf)
```

#### Python

The nycflights13 data is available from the nycflights13 package in Python.
```{python}
from nycflights13 import flights

flights
```

Note there are some differences of this `flights` data from that in tidyverse. The data types for some variables are different. There are no natural ways in Pandas to hold integer column with missing values; so `dep_time` , `arr_time` are `float64` instead of `int64`.  
```{python}
flights.info()
```

To be more consistent with `nycflights13` in tidyverse, we cast `time_hour` to `datetime` type.
```{python}
flights['time_hour'] = pd.to_datetime(flights['time_hour'])
```

#### Julia

Let's use RCall.jl to retrieve the nycflights13 data from R.
```{julia}
using RCall

R"""
library(nycflights13)
"""

flights = rcopy(R"flights")
```

:::


## Create date/times

### `today` and `now`

::: {.panel-tabset}

#### R

Today:
```{r}
# current date
today()
# current date-time
now()
```

#### Python

Today:
```{python}
from datetime import date, datetime

# Current date
date.today()
# Current date-time
datetime.now()
```


#### Julia

```{julia}
using Dates
# Current date
today()
# Current date-time
now()
```

:::

### From strings

::: {.panel-tabset}

#### R

```{r}
ymd("2023-02-02")
mdy("Feb 2nd, 2023")
dmy("02-Feb-2023")
```
```{r}
ymd_hms("2023-02-02 14:57:25")
ymd_hm("2023-02-02 14:57")
```

#### Python

```{python}
from dateutil.parser import parse

datetime.strptime('2023-02-02', '%Y-%m-%d').date()
# I have to use the package python-dateutils to parse Feb 2nd
parse('Feb 2nd, 2023').date()
datetime.strptime('02-Feb-2023', '%d-%b-%Y').date()
```

```{python}
datetime.strptime('2023-02-02 14:57:25', '%Y-%m-%d %H:%M:%S')
datetime.strptime('2023-02-02 14:57', '%Y-%m-%d %H:%M')
```

#### Julia

```{julia}
Date("2023-02-02", dateformat"y-m-d")
# Not sure how to parse "Feb 2nd, 2023" ???
# Date("Feb 2nd, 2023", dateformat"u d, y")
Date("02-Feb-2023", dateformat"d-u-y")
```

```{julia}
DateTime("2023-02-02 14:57:25", dateformat"y-m-d H:M:S")
DateTime("2023-02-02 14:57", dateformat"y-m-d H:M")
```

:::

### From unquoated numbers

::: {.panel-tabset}

#### R

```{r}
ymd(20200202)
```

#### Python

```{python}
datetime(2022, 2, 2).date()
```

#### Julia

```{julia}
Date(2020, 2, 2)
```

:::


### From variables/columns in a tibble

::: {.panel-tabset}

#### R

```{r}
flights |> 
  select(year, month, day, hour, minute) |>
  mutate(departure = make_datetime(year, month, day, hour, minute))
```

```{r}
make_datetime_100 <- function(year, month, day, time) {
  make_datetime(year, month, day, time %/% 100, time %% 100)
}

flights_dt <- flights |> 
  filter(!is.na(dep_time), !is.na(arr_time)) |> 
  mutate(
    dep_time = make_datetime_100(year, month, day, dep_time),
    arr_time = make_datetime_100(year, month, day, arr_time),
    sched_dep_time = make_datetime_100(year, month, day, sched_dep_time),
    sched_arr_time = make_datetime_100(year, month, day, sched_arr_time)
  ) |> 
  select(origin, dest, ends_with("delay"), ends_with("time")) |>
  print(width = Inf)
```
Now we can visualize the distribution of departure times across the year  
```{r}
flights_dt |> 
  ggplot(aes(x= dep_time)) + 
  geom_freqpoly(binwidth = 86400) + # 86400 seconds = 1 day
  labs(
    x = "Departure Time",
    y = "Number of Flights"
  )
```
or within a single day:
```{r}
flights_dt |> 
  filter(dep_time < ymd(20130102)) |> 
  ggplot(aes(dep_time)) + 
  geom_freqpoly(binwidth = 600) # 600 s = 10 minutes
```

#### Python

#### Julia

:::

## Getting components

```{r}
datetime <- ymd_hms("2023-02-02 15:34:56")
year(datetime)
month(datetime)
mday(datetime)
yday(datetime)
wday(datetime)
```
More information in `month()` and `wday()`:
```{r}
month(datetime, label = TRUE, abbr = FALSE)
wday(datetime, label = TRUE, abbr = FALSE)
```
Visualize number of departures during a week:
```{r}
flights_dt |> 
  mutate(wday = wday(dep_time, label = TRUE)) |> 
  ggplot(aes(x = wday)) +
  geom_bar()
```

## Rounding

`floor_date()`, `round_date()`, `ceiling_date()`:
```{r}
flights_dt |> 
  count(week = floor_date(dep_time, "week")) |> 
  ggplot(aes(x = week, y = n)) +
  geom_line()
```

## Time spans

### Durations

Substract two dates we get a `difftime` object:
```{r}
# How old is Hadley?
h_age <- today() - ymd(19791014)
h_age
```
lubridate provides the **duration** object that always uses seconds:
```{r}
as.duration(h_age)
```

Constructors for duration:
```{r}
dseconds(5)
dminutes(10)
dhours(c(12, 24))
ddays(0:5)
dweeks(3)
dyears(1)
```

### Periods

Durations represent an exact number of seconds:
```{r}
one_pm <- ymd_hms("2016-03-12 13:00:00", tz = "America/New_York")
one_pm
one_pm + ddays(1)
```

Periods are time spans but don’t have a fixed length in seconds, instead they work with “human” times, like days and months.
```{r}
one_pm
one_pm + days(1)
```

Constructors for period:
```{r}
seconds(15)
minutes(10)
hours(c(12, 24))
days(7)
months(1:6)
weeks(3)
years(1)
```

Some planes appear to have arrived at their destination before they departed from New York City.
```{r}
flights_dt |> 
  filter(arr_time < dep_time) |>
  print(width = Inf)
```
These are the overnight flights. Let's fix this:
```{r}
flights_dt <- flights_dt |> 
  mutate(
    overnight = arr_time < dep_time,
    arr_time = arr_time + days(overnight * 1),
    sched_arr_time = sched_arr_time + days(overnight * 1)
  ) |>
  print(width = Inf)
```

### Intervals

Intervals are time spans bound by two real date-times. Intervals can be accurately converted to periods and durations.

```{r}
interval(ymd_hms("2009-08-09 13:01:30"), ymd_hms("2009-08-09 12:00:00"))
```

```{r}
date <- ymd_hms("2009-03-08 01:59:59") # DST boundary
date2 <- ymd_hms("2000-02-29 12:00:00")
span <- date2 %--% date
as.duration(span)
as.period(span)
```