---
title: "Visualizing Longitudinal Data Using brolgar"
author: "Dr. Hua Zhou @ UCLA"
date: "Jan 20, 2022"
subtitle: Biostat 203B
output:
  # ioslides_presentation: default
  html_document:
    toc: true
    toc_depth: 4  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Adapted form the tutorial at <http://brolgar.njtierney.com>.

## Longitudinal data

[brolgar](http://brolgar.njtierney.com) is an R package that helps **br**owse **o**ver **l**ongitudinal **d**ata **g**raphically and **a**nalytically in **R**.

## Installing `brolgar`

```{r, eval=F}
install.packages("brolgar")
```

## An example longitudinal data

Load `tidyverse`, `brolgar` and `gghighlight`:
```{r}
library(tidyverse)
library(brolgar)
library(gghighlight)
```

The csv file `wages_pp.csv` contains a data set from the textbook _Applied Longitudinal Data Analysis_ (2003) by Singer and Willett. This data contains measurements on hourly wages by years in the workforce, with education and race as covariates. 
```{bash}
head wages_pp.csv
```

Read following variables into a tibble:  
- `id`: 1-888, for each subject.  
- `lnw`: natural log of wages, adjusted for inflation, to 1990 dollars.  
- `exper`: Experience - the length of time in the workforce (in years). The number of time points and values of time points for each subject can differ.  
- `ged`: when/if a graduate equivalency diploma is obtained.  
- `postexp`: change in experience since getting a ged (if they get one).  
- `black`: categorical indicator of race = black.  
- `hispanic`: categorical indicator of race = hispanic.  
- `hgc`: highest grade completed.  
- `uerate`: unemployment rates in the local geographic region at each measurement time.
```{r}
wages <- read_csv("wages_pp.csv") %>% select(1:8, 10)
wages %>% print(width = Inf)
```

## Turn data frame into a `tsibble`

We turn a regular data frame/tibble into a tidy time series data frame `tsibble` by identifying  
1. The **key** variable is the identifier of individuals.  
2. The **index** variable is the time component of your data.  
3. The **regularity** of the time interveral (index). Longitudinal data typically has irregular time periods between measurements, but can have regular measurements.

```{r}
wages <- wages %>% 
  as_tsibble(key = id, index = exper, regular = FALSE)
wages
```

## Display trajectories of a few individuals

`sample_n_keys()` and `sample_frac_keys()` to take a random sample of keys.
```{r}
set.seed(203)
wages %>%
  sample_n_keys(size = 5) %>%
  ggplot(aes(x = exper,
             y = lnw,
             group = id)) + 
  geom_line()
```

`facet_sample()` allows you to specify the number of keys per facet, and the number of facets with `n_per_facet` and `n_facets`. By default, it splits the data into 12 facets with 3 per facet:
```{r}
set.seed(203)
ggplot(wages,
       aes(x = exper,
           y = lnw,
           group = id)) +
  geom_line() +
  facet_sample()
```

## Find features in longitudinal data

Five number summaries of wages for each individual:
```{r}
wages %>%
  features(lnw,
           feat_five_num)
```
Other feature functions include `feat_monotonic`, `feat_ranges`, `feat_spread`, `feat_three_num`, `n_obs`. For example, find those whose wages only increase or decrease with feat_monotonic:
```{r}
wages %>%
  features(lnw, feat_monotonic)
```

Find the number of observations per individual and summarize by bar plot:
```{r}
wages %>%
  features(lnw, n_obs) %>%
  ggplot(aes(x = n_obs)) + 
    geom_bar()  
```

## Linking individuals back to the data

Once you have created these features, you can join them back to the data with a `left_join`:
```{r}
wages %>%
  features(lnw, feat_monotonic) %>%
  left_join(wages, by = "id") %>%
  ggplot(aes(x = exper,
             y = lnw,
             group = id)) +
  geom_line() + 
  gghighlight(increase)
```
